Why do we Eval?\\\Large A large-scale study of Eval usage in R

### 2021-03-16 analysis/environments.Rmd
Progressing on explaining values

---
title: "Environments and eval in R"
author: "Pierre Donat-Bouillud"
output:
  html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data
  calls_packages_path: /package/summarized.fst
---

>>>>

---
title: "Environments and eval in R"
author: "Pierre Donat-Bouillud"
output:
  html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data/
  calls_packages_path: package/summarized.fst
---

### 2021-03-16 paper/main.tex
Normalized expressions: progress
Progressing on explaining values

\paragraph{Normalized expressions} During tracing, we gather all source expressions and all resolved expression arguments passed to \eval. For instance, a non-resolved expression and some of its resolved expressions can be:

@@

\begin{table}[h]
	\begin{tabular}{|c|c|c|c|c|c|}
		\hline
		Minimized & \#sites & \%sites & \#packages & \#operations & \%envir \\
		\hline
		\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvira \\
		\hline
		\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirb \\
		\hline
		\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirc \\
		\hline
		\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvird \\
		\hline
		\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvire \\
		\hline
		\texttt{\$}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirf \\
		\hline
		\texttt{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirg \\
		\hline
		\texttt{<-} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirh \\
		\hline
		\texttt{FUN} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenviri \\
		\hline
		\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirj \\
		\hline
	\end{tabular}
	\caption{Minimized expressions: ten most frequent in terms of sites. \#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites); \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is. \#operations is the average number of steps performed by the R interpreter when given that minimized expression, and \%envir is the percentage of call sites for which the \c{envir} argument is explicitly passed.} \label{tab:minimizedexpressions}
\end{table}

The most common minimized expression is a variable.\footnote{An arithmetic with some variables and constant will also be normalized to \c{X}, i.e. a variable.} It happens in \packageMinimizedpropsitesa of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus. The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites, which suggests that \eval is used here to get the value of the variable in a non-local environment.

Compositions of functions, without variables, is the second most frequent minimized expression.

When a function is evaluated on some variable, the percentage of non default environment is high.

\c{FUN} represents a function definition

\c{{BLOCK}} is a sequence of statements. It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.

@@

Note that a site can exhibit several minimized expressions. However, \packageNbOneMinimizedPercent of the sites have only one minimized expression. The most common combinations are:

\NOTE{Maybe we should ditch the table of the minimized expressions and rather use the ones of combinations. It has the advantage of showing what is the most frequent, per call site. But it's a long tail, and sites that have several minimized expressions are not that numerous, only 13\%}

>>>>

\subsection{Normalized expressions} During tracing, we gather all source expressions and all resolved expression arguments passed to \eval. For instance, a non-resolved expression and some of its resolved expressions can be:

@@

\begin{table}[h]
	\begin{tabular}{|c|c|c|c|c|c|}
		\hline
		Minimized & \#sites & \%sites & \#packages & \#operations & \%envir \\
		\hline
		\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvira \\
		\hline
		\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirb \\
		\hline
		\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirc \\
		\hline
		\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvird \\
		\hline
		\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvire \\
		\hline
		\texttt{\$}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirf \\
		\hline
		\texttt{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirg \\
		\hline
		\texttt{<-} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirh \\
		\hline
		\texttt{FUN} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenviri \\
		\hline
		\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirj \\
		\hline
	\end{tabular}
	\caption{Minimized expressions: ten most frequent in terms of sites. \#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites) with at least this minimized expression; \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is. \#operations is the average number of steps performed by the R interpreter when given that minimized expression, and \%envir is the percentage of call sites for which the \c{envir} argument is explicitly passed.} \label{tab:minimizedexpressions}
\end{table}

Note that a site can exhibit several minimized expressions. However, \packageNbOneMinimizedPercent of the sites have only one minimized expression.

\paragraph{Variables.}
The most common minimized expression is a variable.\footnote{An arithmetic with some variables and constant will also be normalized to \c{X}, i.e. a variable.} It happens in \packageMinimizedpropsitesa of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus. The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites, which suggests that \eval is used here to get the value of the variable in a non-local environment.

\paragraph{Composition of functions. } Compositions of functions, without variables, is the second most frequent minimized expression.

\paragraph{Value.}
Many sites evaluates to \c{V}.  \packageValOneNodePercent of those sites are only passed an expression with one node, \ie one value, not a more complex expression that is later constant folded. \c{V} corresponds to two main categories of expressions passed to \eval:

\begin{itemize}
	\item \eval is passed an actual value, a double, an integer, an environment, a closure and so on
	\item \eval is passed a language expression that trivially evaluates to a value, for instance, \c{quote(1)}, \quote{c("a", "b", "c")}, and does not include variables, so does not depend on a specific environment.
\end{itemize}

For the first category, \eval is not necessary. In \packageNbCallSitesUniqueActualValue sites, \ie \packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual values. The median number of runs for those sites is \packageMedianRunSitesUniqueActualValue, suggesting it happens because there are rarely triggered in our dataset and all paths are not taken.

The percentage of non default environment passed to \eval is \packageMinimizedpercentenvirc. Some sites are passed both a value or a symbol, which explains that a specific environment is precised.

%TODO: find out the percentage of non-default envir for sites which have both a value and a variable
%TODO: find the same percentage for sites which have at least one variable passed to them?

When a function is evaluated on some variable (\c{F(X)}), the percentage of non default environment is high, \packageMinimizedpercentenvird\%.

\c{FUN} represents a function definition. It entails very few steps of the interpreter, as the body of the new function is just stored for later execution.

\c{\{BLOCK\}} is a sequence of statements. It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.

@@

The most common combinations for sites with several minimized expressions are:

\NOTE{Maybe we should ditch the table of the minimized expressions and rather use the ones of combinations. It has the advantage of showing what is the most frequent, per call site. But it's a long tail, and sites that have several minimized expressions are not that numerous, only 13\%. But if include combinations, percentages change. For instance, 3314 call sites have only X, i.e. 22.51, instead of 4353 and 29.6\%. }

\NOTE{Two small paragraphs about normalized expressions in Base and Kaggle.}
